---
title: "mIF SSL - Component 6 DE"
format: html
editor: visual
execute: 
  warning: false
---

## Imports and preprocessing

### For RNA-Seq

```{r}
#| code-fold: true
#| code-summary: "Code"
#| echo: false
library(tidyverse)

component_data <- read_csv('/Users/Kai/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/PhD_Workspace/Hyperstack_paper/nmf/W_df_k12.csv')

load("/Users/Kai/Library/CloudStorage/OneDrive-UniversityofGlasgow/Temposeq/LUADChohort.RData")
raw_counts <- LUADCohort$rawExpr

metadata <- LUADCohort$Metadata %>% 
  mutate(samples = sprintf("ACA_%04s", `Case Number`)) %>% 
  right_join(component_data, join_by(samples == samples)) %>% 
  dplyr:::select(c("...1", "Component_1", "Component_2", "Component_3", "Component_4", "Component_5", 
                   "Component_6", "Component_7", "Component_8", "Component_9", "Component_10", 
                   "Component_11", "Component_12", "samples")) %>% 
  na.omit() %>% 
  column_to_rownames(var = "...1")

hallmarks_ssgsea <- read_csv("/Users/Kai/Library/CloudStorage/OneDrive-UniversityofGlasgow/Temposeq/final_figures/v250331_superclusters/all_LATTICeA_ssGSEA_hallmarks_norm.csv") %>% 
  column_to_rownames("pathway") %>% 
  dplyr:::select(rownames(metadata))
```

n=784 cores with paired RNASeq vs. whole patient NMF scoring

### For survival analysis

```{r}
#| echo: false
comp_survival <- read_csv("/Users/Kai/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/PhD_Workspace/Hyperstack_paper/nmf/W_df_k12_survival.csv") %>% 
  dplyr::select(-c("...1")) %>% 
  column_to_rownames("samples")

```

n=906 non-mucinous adenocarcinoma with survival data.

## Survival analysis

### Split by component 6 score

```{r}
#| code-fold: true
#| code-summary: "Code to generate the split"
library(survminer)
comp_6_cut <- surv_cutpoint(comp_survival,
                             time = "os_event_data",
                             event = "os_event_ind",
                             variables = "Component_6")

summary(comp_6_cut)
```

```{r}
#| code-fold: true
#| code-summary: "Code for KM"
#| fig-cap: "Survival function for patients with high component 6 scores. High n=342, Low n=564"
library(survival)
comp6_classified_df <- surv_categorize(comp_6_cut) 
km_fit <- survfit(Surv(os_event_data, os_event_ind) ~ Component_6, data = comp6_classified_df)
surv_diff <- survdiff(Surv(os_event_data, os_event_ind) ~ Component_6, data = comp6_classified_df)
comp6_classified_df$samples = rownames(comp_survival)

ggsurvplot(km_fit, 
           pval = paste0("p = ", sprintf("%.4f", surv_diff$pvalue)),
           pvalue.format = TRUE,
           xlim = c(0, 60),
           break.x.by = 10,
           legend = "top",
           legend.title = "Component 6",
           legend.labs = c("High", "Low"),
           data = comp6_classified_df) + 
  labs(x = "Time (months)", y = "Survival probability (overall survival)", title = "All patients")

ggsave("KM_comp6_all.pdf",
       path = "~/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/Collaborations/mIF SSL/",
       scale = 0.5,
       width = 8,
       height = 8,
       dpi = 300
)
```

### Sub-select by grade 1/2

```{r}
#| code-fold: true
#| code-summary: "Code for KM"
#| fig-cap: "Survival function for patients (whole tumour grade 1/2 only) with high component 6 scores. High n=110, Low n=274"

comp_survival_lowgrade <- comp_survival %>% 
  filter(G3 != 1)

comp_6_cut_lowgrade <- surv_cutpoint(comp_survival_lowgrade,
                             time = "os_event_data",
                             event = "os_event_ind",
                             variables = "Component_6")


comp6_classified_df_lg <- surv_categorize(comp_6_cut_lowgrade) 
km_fit <- survfit(Surv(os_event_data, os_event_ind) ~ Component_6, data = comp6_classified_df_lg)
surv_diff <- survdiff(Surv(os_event_data, os_event_ind) ~ Component_6, data = comp6_classified_df_lg)

ggsurvplot(km_fit, 
           pval = paste0("p = ", sprintf("%.4f", surv_diff$pvalue)),
           pvalue.format = TRUE,
           xlim = c(0, 60),
           break.x.by = 10,
           legend = "top",
           legend.title = "Component 6",
           legend.labs = c("High", "Low"),
           data = comp6_classified_df_lg) + 
  labs(x = "Time (months)", y = "Survival probability (overall survival)", title = "Grade 1/2 only")

ggsave("KM_comp6_LG_only.pdf",
       path = "~/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/Collaborations/mIF SSL/",
       scale = 0.5,
       width = 8,
       height = 8,
       dpi = 300
)
```

### Sub-select by grade 3

```{r}
#| code-fold: true
#| code-summary: "Code for KM"
#| fig-cap: "Survival function for patients (whole tumour grade 3 only) with high component 6 scores. High n=136, Low n=386"
comp_survival_highgrade <- comp_survival %>% 
  filter(G3 == 1)

comp_6_cut_highgrade <- surv_cutpoint(comp_survival_highgrade,
                             time = "os_event_data",
                             event = "os_event_ind",
                             variables = "Component_6")


comp6_classified_df_hg <- surv_categorize(comp_6_cut_highgrade) 
km_fit <- survfit(Surv(os_event_data, os_event_ind) ~ Component_6, data = comp6_classified_df_hg)
surv_diff <- survdiff(Surv(os_event_data, os_event_ind) ~ Component_6, data = comp6_classified_df_hg)

ggsurvplot(km_fit, 
           pval = paste0("p = ", sprintf("%.4f", surv_diff$pvalue)),
           pvalue.format = TRUE,
           xlim = c(0, 60),
           break.x.by = 10,
           legend = "top",
           legend.title = "Component 6",
           legend.labs = c("High", "Low"),
           data = comp6_classified_df_hg) + 
  labs(x = "Time (months)", y = "Survival probability (overall survival)", title = "Grade 3 only")

ggsave("KM_comp6_HG_only.pdf",
       path = "~/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/Collaborations/mIF SSL/",
       scale = 0.5,
       width = 8,
       height = 8,
       dpi = 300
)

```

## RNA-Seq analysis

### Differential expression by survival split

```{r}
#| echo: false
library(DESeq2)

metadata_comp6 <- metadata %>% 
  rownames_to_column(var = "ts_id") %>% 
  left_join(comp6_classified_df, join_by(samples == samples)) %>% 
  column_to_rownames(var = "ts_id") %>% 
  na.omit() %>% 
  rename(Component_6.y = "comp6_strata")

metadata_comp6$comp6_strata <- factor(metadata_comp6$comp6_strata)

filtered_raw_counts <- as.data.frame(raw_counts) %>% 
  select(rownames(metadata_comp6))

raw_counts_filtered_genes <- round(filtered_raw_counts, 0) %>%
  subset(., apply(filtered_raw_counts, 1, mean) >= 10) %>% 
  na.omit(.)

raw_counts_matrix <- raw_counts_filtered_genes %>% 
  dplyr:::select(rownames(metadata_comp6)) %>% 
  as.matrix(.)

metadata_comp6 <- metadata_comp6[colnames(filtered_raw_counts), ]

dds <- DESeqDataSetFromMatrix(countData = raw_counts_matrix, colData = metadata_comp6, design = ~ comp6_strata)
dds <- DESeq(dds,
             BPPARAM = MulticoreParam(workers = detectCores() - 1))  

resdata <- data.frame(round(counts(dds, normalized = TRUE), 2))
colnames(resdata) <- colnames(raw_counts_matrix)

de <- data.frame(results(dds, c("comp6_strata", "high", "low"))) %>% 
  arrange(padj)

de$mlog10p <- -log(de$padj, 10)

sig_up <- de %>% 
  filter(log2FoldChange > 1 & padj < 0.01)

sig_down <- de %>% 
  filter(log2FoldChange < -1 & padj < 0.01)

```

```{r}
library(ggrepel)
ggplot(de, aes(x = log2FoldChange, y = mlog10p)) +
  geom_point(aes(color = "Not Significant")) +
  geom_point(data = sig_up, aes(color = "Up in Component 6")) +
  geom_point(data = sig_down, aes(color = "Down in Component 6")) +
  scale_color_manual(values = c(
    "Not Significant" = "grey",
    "Up in Component 6" = "#E15759",
    "Down in Component 6" = "#4E79A7"
  ), name = "") +
  geom_text_repel(data = sig_up %>% slice_head(n = 5), aes(label = row.names(sig_up %>% slice_head(n = 5))), max.overlaps = 40) + 
  geom_text_repel(data = sig_down %>% slice_head(n = 5), aes(label = row.names(sig_down %>% slice_head(n = 5))), max.overlaps = 40) + 
  labs(x = 'L2FC', y = '-log10p', title = 'DE Component 6') +
  theme_bw() + 
  theme(panel.grid = element_blank())

ggsave("DE_comp6_volcano.pdf",
       path = "~/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/Collaborations/mIF SSL/",
       scale = 0.5,
       width = 12,
       height = 8,
       dpi = 300
)
```

#### GSEA: Hallmarks

```{r}
library(fgsea)
library(parallel)
library(BiocParallel)
library(exCITingpath)
library(scales)

# Rank genes for GSEA (L2FC)
ranked_genes <- de$log2FoldChange
names(ranked_genes) <- row.names(de)
ranked_genes <- sort(ranked_genes, decreasing = TRUE)

# Hallmarks

hallmarks <- loadDB("h.all.v2024.1.Hs.symbols.gmt")

gsea_results_hallmarks <- fgsea(
  pathways = hallmarks,
  stats = ranked_genes,      
  minSize = 5,
  maxSize = length(ranked_genes) - 1,
  BPPARAM = MulticoreParam(workers = detectCores() - 1)
)

top_pathways <- filter(gsea_results_hallmarks, padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  mutate(pathway = sub("^HALLMARK_", "", pathway)) %>% 
  mutate(pathway = str_wrap(pathway, width = 50)) %>% 
  slice_head(n = 20)

top_pathways$mlog10p <- -log10(top_pathways$padj)

ggplot(top_pathways, aes(x = reorder(pathway, NES), y = NES, fill = mlog10p)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Up in Component 6 High", x = "Pathways", y = "Normalised Enrichment Score") +
  scale_fill_gradient(high = "#E15759", low = "gray", name = "-log10(p)", limits = c(2, 10), oob=squish) +
  theme_bw() + 
  theme(panel.grid = element_blank())

ggsave("DE_comp6_GSEA_hallmarks.pdf",
       path = "~/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/Collaborations/mIF SSL/",
       scale = 0.5,
       width = 12,
       height = 8,
       dpi = 300
)
```

#### GSEA: Kegg

```{r}
kegg <- loadDB("KEGG_2021_Human.txt")

gsea_results_kegg <- fgsea(
  pathways = kegg,
  stats = ranked_genes,      
  minSize = 5,
  maxSize = length(ranked_genes) - 1,
  BPPARAM = MulticoreParam(workers = detectCores() - 1)
)

top_pathways <- filter(gsea_results_kegg, padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  mutate(pathway = str_wrap(pathway, width = 50)) %>% 
  slice_head(n = 20)

top_pathways$mlog10p <- -log10(top_pathways$padj)

ggplot(top_pathways, aes(x = reorder(pathway, NES), y = NES, fill = mlog10p)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Up in Component 6 High", x = "Pathways", y = "Normalised Enrichment Score") +
  scale_fill_gradient(high = "#E15759", low = "gray", name = "-log10(p)", limits = c(2, 10), oob=squish) +
  theme_bw() + 
  theme(panel.grid = element_blank())

ggsave("DE_comp6_GSEA_kegg.pdf",
       path = "~/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/Collaborations/mIF SSL/",
       scale = 0.5,
       width = 12,
       height = 8,
       dpi = 300
)
```

#### GSEA: GO

```{r}
go_all <- loadDB(c("GO_Biological_Process_2023.txt", "GO_Cellular_Component_2023.txt", "GO_Molecular_Function_2023.txt"))

gsea_results_go <- fgsea(
  pathways = go_all,
  stats = ranked_genes,      
  minSize = 5,
  maxSize = length(ranked_genes) - 1,
  BPPARAM = MulticoreParam(workers = detectCores() - 1)
)

top_pathways <- filter(gsea_results_go, padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  mutate(pathway = gsub("\\(.*?\\)", "", pathway)) %>% 
  mutate(pathway = str_wrap(pathway, width = 50))

top_pathways$mlog10p <- -log10(top_pathways$padj)

ggplot(top_pathways, aes(x = reorder(pathway, NES), y = NES, fill = mlog10p)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Up in Component 6 High", x = "Pathways", y = "Normalised Enrichment Score") +
  scale_fill_gradient(high = "#E15759", low = "gray", name = "-log10(p)", limits = c(2, 10), oob=squish) +
  theme_bw() + 
  theme(panel.grid = element_blank())
```

#### ORA: GO

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

de_genes <- de %>% 
  rownames_to_column("gene") %>% 
  filter(log2FoldChange > 1 & padj < 0.05) %>% 
  pull(gene)

enrich_go_bp <- enrichGO(
  gene = de_genes,
  universe = rownames(resdata),
  keyType = "SYMBOL",
  OrgDb = org.Hs.eg.db,
  ont = "ALL",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.1,
  readable = TRUE)

dotplot(enrich_go_bp)

```

#### ORA: KEGG

```{r}
de_df <- bitr(de_genes, fromType = "SYMBOL",
                toType = c("ENTREZID"),
                OrgDb = org.Hs.eg.db)

ora_kegg <- enrichKEGG(
  gene = de_df$ENTREZID,
  organism = "hsa",
  universe = rownames(resdata),
  pvalueCutoff = 0.05,
  minGSSize = 1)

```

### Differential expression in high grade tumours

```{r}
#| echo: false

metadata_comp6_hg <- metadata %>% 
  rownames_to_column(var = "ts_id") %>% 
  left_join(comp6_classified_df_hg %>% rownames_to_column("samples"), join_by(samples == samples)) %>% 
  column_to_rownames(var = "ts_id") %>% 
  na.omit() %>% 
  rename(Component_6.y = "comp6_strata")

metadata_comp6_hg$comp6_strata <- factor(metadata_comp6_hg$comp6_strata)

filtered_raw_counts <- as.data.frame(raw_counts) %>% 
  select(rownames(metadata_comp6_hg))

raw_counts_filtered_genes <- round(filtered_raw_counts, 0) %>%
  subset(., apply(filtered_raw_counts, 1, mean) >= 10) %>% 
  na.omit(.)

raw_counts_matrix <- raw_counts_filtered_genes %>% 
  dplyr:::select(rownames(metadata_comp6_hg)) %>% 
  as.matrix(.)

metadata_comp6_hg <- metadata_comp6_hg[colnames(filtered_raw_counts), ]

dds <- DESeqDataSetFromMatrix(countData = raw_counts_matrix, colData = metadata_comp6_hg, design = ~ comp6_strata)
dds <- DESeq(dds,
             BPPARAM = MulticoreParam(workers = detectCores() - 1))  

resdata <- data.frame(round(counts(dds, normalized = TRUE), 2))
colnames(resdata) <- colnames(raw_counts_matrix)

de <- data.frame(results(dds, c("comp6_strata", "high", "low"))) %>% 
  arrange(padj)

de$mlog10p <- -log(de$padj, 10)

sig_up <- de %>% 
  filter(log2FoldChange > 1 & padj < 0.01)

sig_down <- de %>% 
  filter(log2FoldChange < -1 & padj < 0.01)
```

```{r}
#| fig-cap: "Survival function for patients (whole tumour grade 3 only) with high component 6 scores. High n=109, Low n=305"
ggplot(de, aes(x = log2FoldChange, y = mlog10p)) +
  geom_point(aes(color = "Not Significant")) +
  geom_point(data = sig_up, aes(color = "Up in Component 6 High")) +
  geom_point(data = sig_down, aes(color = "Down in Component 6 High")) +
  scale_color_manual(values = c(
    "Not Significant" = "grey",
    "Up in Component 6 High" = "#E15759",
    "Down in Component 6 High" = "#4E79A7"
  ), name = "") +
  geom_text_repel(data = sig_up %>% slice_head(n = 10), aes(label = row.names(sig_up %>% slice_head(n = 10))), max.overlaps = 40) + 
  geom_text_repel(data = sig_down %>% slice_head(n = 10), aes(label = row.names(sig_down %>% slice_head(n = 10))), max.overlaps = 40) + 
  labs(x = 'L2FC', y = '-log10p', title = 'DE Component 6') +
  theme_bw() + 
  theme(panel.grid = element_blank())
```

#### GSEA: Hallmarks

```{r}
# Rank genes for GSEA (L2FC)

ranked_genes <- de$log2FoldChange
names(ranked_genes) <- row.names(de)
ranked_genes <- sort(ranked_genes, decreasing = TRUE)

gsea_results_hallmarks <- fgsea(
  pathways = hallmarks,
  stats = ranked_genes,      
  minSize = 5,
  maxSize = length(ranked_genes) - 1,
  BPPARAM = MulticoreParam(workers = detectCores() - 1)
)

top_pathways <- filter(gsea_results_hallmarks, padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  mutate(pathway = sub("^HALLMARK_", "", pathway)) %>% 
  mutate(pathway = str_wrap(pathway, width = 50)) %>% 
  slice_head(n = 20)

top_pathways$mlog10p <- -log10(top_pathways$padj)

ggplot(top_pathways, aes(x = reorder(pathway, NES), y = NES, fill = mlog10p)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Up in Component 6 High", x = "Pathways", y = "Normalised Enrichment Score") +
  scale_fill_gradient(high = "#E15759", low = "gray", name = "-log10(p)", limits = c(2, 10), oob=squish) +
  theme_bw() + 
  theme(panel.grid = element_blank())
```

#### GSEA: KEGG

```{r}
gsea_results_kegg <- fgsea(
  pathways = kegg,
  stats = ranked_genes,      
  minSize = 5,
  maxSize = length(ranked_genes) - 1,
  BPPARAM = MulticoreParam(workers = detectCores() - 1)
)

top_pathways <- filter(gsea_results_kegg, padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  mutate(pathway = str_wrap(pathway, width = 50)) %>% 
  slice_head(n = 20)

top_pathways$mlog10p <- -log10(top_pathways$padj)

ggplot(top_pathways, aes(x = reorder(pathway, NES), y = NES, fill = mlog10p)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Up in Component 6 High", x = "Pathways", y = "Normalised Enrichment Score") +
  scale_fill_gradient(high = "#E15759", low = "gray", name = "-log10(p)", limits = c(2, 10), oob=squish) +
  theme_bw() + 
  theme(panel.grid = element_blank())
```

#### GSEA: GO

```{r}
gsea_results_go <- fgsea(
  pathways = go_all,
  stats = ranked_genes,      
  minSize = 5,
  maxSize = length(ranked_genes) - 1,
  BPPARAM = MulticoreParam(workers = detectCores() - 1)
)

top_pathways <- filter(gsea_results_go, padj < 0.05) %>%
  arrange(desc(abs(NES))) %>%
  mutate(pathway = gsub("\\(.*?\\)", "", pathway)) %>% 
  mutate(pathway = str_wrap(pathway, width = 50))

top_pathways$mlog10p <- -log10(top_pathways$padj)

ggplot(top_pathways, aes(x = reorder(pathway, NES), y = NES, fill = mlog10p)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Up in Component 6 High", x = "Pathways", y = "Normalised Enrichment Score") +
  scale_fill_gradient(high = "#E15759", low = "gray", name = "-log10(p)", limits = c(2, 10), oob=squish) +
  theme_bw() + 
  theme(panel.grid = element_blank())
```

### ssGSEA vs. all components

#### Prepare correlation matrix

```{r}
library(ComplexHeatmap)
library(circlize)
compute_correlations_with_pvals <- function(matrix1, matrix2, corr_method) {
  # Initialize matrices for correlations and p-values
  n_rows <- ncol(matrix1)
  n_cols <- ncol(matrix2)
  cors <- matrix(NA, nrow = n_rows, ncol = n_cols)
  pvals <- matrix(NA, nrow = n_rows, ncol = n_cols)
  
  # Compute correlations and p-values
  for (i in 1:n_rows) {
    for (j in 1:n_cols) {
      test_result <- cor.test(matrix1[,i], matrix2[,j], method = corr_method)
      cors[i,j] <- test_result$estimate
      pvals[i,j] <- test_result$p.value
    }
  }
  
  # Adjust p-values for multiple testing
  # adjusted_pvals <- matrix(p.adjust(pvals, method = "BH"), nrow = n_rows)
  adjusted_pvals <- matrix(p.adjust(pvals, method = "BH"), nrow = n_rows, ncol = n_cols)
  
  return(list(correlations = cors, 
              raw_pvals = pvals, 
              adjusted_pvals = adjusted_pvals))
}

stat_results <- compute_correlations_with_pvals(t(hallmarks_ssgsea), metadata %>% dplyr:::select(-samples), corr_method = "spearman")

stat_cor <- stat_results$correlations
rownames(stat_cor) <- gsub("HALLMARK_", "", rownames(hallmarks_ssgsea))
colnames(stat_cor) <- sub("_", " ", colnames(metadata %>% dplyr:::select(-samples)))

sig_matrix <- matrix("", nrow = nrow(stat_results$adjusted_pvals), ncol = ncol(stat_results$adjusted_pvals))
sig_matrix[stat_results$adjusted_pvals < 0.01] <- "*"

# sig_matrix[stat_results$adjusted_pvals < 0.01 & stat_results$adjusted_pvals >= 0.001] <- "**"
# sig_matrix[stat_results$adjusted_pvals < 0.05 & stat_results$adjusted_pvals >= 0.01] <- "*"

```

#### Draw heatmap

```{r}
col_colors <- colorRamp2(c(-0.4, 0, 0.4), c("#4E79A7", "white", "#E15759"))

p <- Heatmap(
  stat_cor,
  name = "Spearman\ncorrelation",
  col = col_colors,
  border = TRUE,
  row_names_side = "left",
  row_dend_side = "right", 
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (sig_matrix[i, j] != "") {
      grid.text(sig_matrix[i, j], x, y)
    }
  }
)

pdf(file="/Users/Kai/Library/CloudStorage/OneDrive-UniversityofGlasgow/Documents/Collaborations/mIF\ SSL/ssGSEA_vs_NMF12-stats.pdf", width = 7.2, height = 9)
draw(p, heatmap_legend_side = "right", annotation_legend_side = "right", padding = unit(c(2, 30, 2, 2), "mm"))
dev.off()
```